---
title: "R Tip of the Day"
subtitle: "Interactive Maps with {leaflet}"
author: "Willemijn Leiner"
format:
  revealjs:
    theme: moon
    slide-number: true
    chalkboard: true
editor: visual
---

```{r}
library(sf)
library(leaflet)
library(dplyr)
```

## What is {leaflet}?

- `{leaflet}` lets you build **interactive maps in R**
- Based on the **Leaflet JavaScript library**, which also powers many web pages 
- with the `{leaflet}` package you can do some cool things: 
- display geographical data!
- add markers and shapes
- interactive popups 

## What is it useful for?

- Interactive maps are useful for a lot of things: 
  - city planning
  - environmental science
  - visualization in research
  - transportation data

- Personal Interest in maps!
  - Reminds me of GIS mapping
  - TimeGuesser
  - I grew up in **Utrecht in the Netherlands**


## Basic Map of the World

```{.r code-line-numbers="1|2"}
leaflet() %>%
  addProviderTiles("CartoDB.Positron")
```

- **leaflet()** initializes a new interactive map
- **%>%** passes the map object to the next function
- **addProviderTiles("CartoDB.Positron")** adds the background map layer

## Basic Map of the World

```{r}
leaflet() %>%
  addProviderTiles("CartoDB.Positron")
```

## Zooming in on a Location

```{.r code-line-numbers="1-2|4|6|7-10"}
utrecht_lat <- 52.0907
utrecht_lng <- 5.1214

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(lng = utrecht_lng, lat = utrecht_lat, zoom = 13) %>%
  addMarkers(
    lng = utrecht_lng,
    lat = utrecht_lat,
    popup = "Utrecht"
  )
```

- store **latitude and longitude** of Utrecht
- **leaflet()** creates this interactive map 
- **setView()** centers the map and controls the zoom level
- **addMarkers()** adds a clickable popup marker 

## Zooming in on a Location

```{r}
utrecht_lat <- 52.0907
utrecht_lng <- 5.1214

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  setView(lng = utrecht_lng, lat = utrecht_lat, zoom = 13) %>%
  addMarkers(lng = utrecht_lng,
             lat = utrecht_lat,
             popup = "Utrecht")
```

## The best places to be in Utrecht!

```{.r code-line-numbers="1-12|13-21"}
bike_spots <- data.frame(
  name = c(
    "Utrecht Centraal – Jaarbeurszijde",
    "De Dom Toren (The Dumb Tower)",
    "Atletiekbaan Maarschalkerweerd - U-Track Club",
    "My House!",
    "Favorite Place!",
    "Wilhelmina Park"
  ),
  lat = c(52.0894, 52.090663, 52.078978, 52.085239, 52.092751, 52.087659),
  lng = c(5.1100, 5.121204, 5.156714, 5.147264, 5.12208, 5.140606),

  capacity = c(12500, 6000, 50, 4, 30, 250),

  image = c(
    "",
    "https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/unnamed.jpg",
    "https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/IMG_0416.jpeg",
    "https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/my_house.jpeg",
    "https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/favoriete_restaurant.jpeg",
    "https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/wilhelmina_park.jpeg"
  )
)
```

- First create data frame: each row represents one location
- Store coordinates using latitude and longitude + extra information

## The best places to be in Utrecht!

```{.r code-line-numbers="1|2|3"}
leaflet(bike_spots) %>%
  addProviderTiles("OpenStreetMap") %>%
  setView(lng = 5.1214, lat = 52.0907, zoom = 13)
```

- **leaflet(bike_spots)** to clarify data
- **addProviderTiles()** for the background map
- **setView()** centers the map 

## The best places to be in Utrecht!

```{.r code-line-numbers="1-5|6-11"}
  addCircleMarkers(
    ~lng, ~lat,
    radius = 5,
    color = "red",
    fillOpacity = 0.9,
    popup = ~paste0(
      "<b>", name, "</b><br>",
      "Capacity: ", capacity, " bikes<br>",
      "<img src='", image, "' width='200px'>"
    )
  )
```

- **addCircleMarkers()** and **~lng, ~lat, ** tell leaflet to add markers to the locations I have earlier mapped out
- **popup** allows HTML content which are the popups. 

## The best places to be in Utrecht!

```{r}

library(leaflet)

bike_spots <- data.frame(
  name = c(
    "Utrecht Centraal – Jaarbeurszijde",
    "De Dom Toren (The Dumb Tower)",
    "Atletiekbaan Maarschalkerweerd - U-Track Club",
    "My House!", 
    "Favorite PLace!", 
    "Wilhelmina Park"
  ),
  lat = c(52.0894, 52.090663, 52.078978, 52.085239, 52.092751, 52.087659),
  lng = c(5.1100, 5.121204, 5.156714, 5.147264, 5.12208, 5.140606),
  capacity = c(12500, 6000, 50, 4, 30, 250),
  image = c(
  "", 
  'https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/unnamed.jpg', 
  'https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/IMG_0416.jpeg', 
  'https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/my_house.jpeg',
  'https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/favoriete_restaurant.jpeg', 
  'https://raw.githubusercontent.com/leiner3/SDS264_spring2026/main/wilhelmina_park.jpeg
'
  )
)

leaflet(bike_spots) %>%
  addProviderTiles("OpenStreetMap") %>% # Here I added the provider tiles "OpenStreetMap", which is the most descriptive map of all of the other tiles I have tried. 
  setView(lng = 5.1214, lat = 52.0907, zoom = 13) %>%
  addCircleMarkers(
    ~lng, ~lat,
    radius = 4,
    color = "red",
    fillOpacity = 0.9,
    popup = ~paste0(
      "<b>", name, "</b><br>",
      "Capacity: ", capacity, " bikes<br>",
      "<img src='", image, "' width='200px'>"
    )
  )
```

